--- 
title: langClass package
format: gfm
---
# langclass

## A Language classifier using Naive Bayes

Identify the language of a text using the ultra speed of a Naive Bayes
classifier.

This package uses [Quanteda::stopwords](https://github.com/quanteda/stopwords)
(see the [documentation](https://stopwords.quanteda.io/reference/index.html))
to train a Naive Bayes classifier over stopword lists for all languages
available in the `stopwords` package, then classifies new texts into one of
those languages.

## Installation

```r
# install.packages("remotes")
remotes::install_github("SoaresAlisson/langclass")
```

## Usage

```{r ex}
library(langclass)

available_languages()

# Classify texts right away with the shipped pre-trained classifier
classify_text("Olá pessoal, meu nome é João")
classify_text(c("Hello everyone, my name is John",
                "Ciao a tutti, il mio nome è Marco"))
```

Or build your own classifier and use it
```{r build_classifier}
#| eval: false
model <- build_classifier()
classify_text("Bonjour tout le monde, mon nom est Anne", model = model)
```

## Functions

- `available_languages()`: list all languages covered by the package.
- `build_classifier()`: (re)build a Naive Bayes classifier from the
  stopword lists.
- `classify_text()`: classify one or more texts into a language, returning
  the best match (and optionally the full probability table).
