---
title: "Classifying texts by language with Naive Bayes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classifying texts by language with Naive Bayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

`langclass` identifies the language of a text using a Naive Bayes classifier
trained on the stopword lists provided by the
[`stopwords`](https://github.com/quanteda/stopwords) package. Because a
language's stopwords are highly characteristic of that language, a classifier
built over them is fast and accurate for common languages, and it ships with
the package so you can classify texts with no extra setup.

## The classifier

The model is trained as follows:

1. For every language code reported by `available_languages()`, we fetch that
   language's stopword list.
2. Each list becomes a "document" whose class label is the language code.
3. A Naive Bayes model is fitted with
   `quanteda.textmodels::textmodel_nb()` using a uniform prior.

To classify a new text, it is tokenized and lower-cased, aligned to the
features seen during training, and the posterior probability of each language
is computed.

## Listing the available languages

```{r}
library(langclass)

# All languages covered by the default classifier
langs <- available_languages()
head(langs)
length(langs)
```

By default we use the `stopwords-iso` source, which has the widest coverage.
You can ask for a different source:

```{r}
available_languages(source = "snowball")
```

## Classifying a single text

```{r}
classify_text("Olá pessoal, meu nome é João")
```

The result is a named character vector with the most likely language code.

## Classifying many texts

`classify_text()` is vectorised, so you can pass many texts at once:

```{r}
texts <- c(
  "Hello everyone, my name is John",
  "Hallo zusammen, ich bin der Lukas",
  "Hola a todos, soy Lucía",
  "Bonjour tout le mon, je m'appelle Louise",
  "Ciao a tutti, mi chiamo Sofia"
)

classify_text(texts)
```

## Getting the full probability table

Pass `details = TRUE` to obtain the posterior probability for every language,
plus a `label` and `probability` column:

```{r}
probs <- classify_text("Bonjour tout le mon, je m'appelle Louise",
                       details = TRUE)
probs[, c("label", "probability", "fr", "it", "pt")]
```

## Handling low-confidence predictions

Some texts (numbers, nonsense, very short strings) contain almost no stopword
signal and will score close to the uniform prior. Set `threshold` to return
`NA` for any text whose best probability falls below it:

```{r}
classify_text("123", threshold = 0.05)
classify_text("xwa asd lorem ipsum", threshold = 0.05)
classify_text("Hello everyone, my name is John", threshold = 0.05)
```

## Building your own classifier

The package ships a pre-trained model covering all `stopwords-iso` languages,
which is what `classify_text()` uses by default. You can also build a model
yourself — for example, restricting to a single source such as `snowball` —
and pass it explicitly:

```{r}
model <- build_classifier(source = "snowball")
classify_text("Hallo Leute, mein Name ist Johannes", model = model)
```

## How it works

The key idea is that each language contributes exactly one training document
containing its stopwords. Consider the German and English lists: they overlap
in a handful of words but differ sharply in the rest (`und`, `der`, `ich`,
`nicht` vs. `the`, `and`, `you`, `not`). The Naive Bayes classifier learns
these contrasts, so a text dense in German stopwords strongly favours German.

The resulting model is small (about 80 KB) because it only tracks the
presence of stopword features per language, which is why `langclass` can
ship it inside the package and classify instantly.

## Limitations

`langclass` is a fast heuristic, not a production-grade detector. The model
is trained on stopword lists alone — one small "document" per language —
and content words are ignored during classification, so it has real limits:

- **It only sees stopwords.** Texts with few or no stopwords carry almost
  no signal, and the prediction is essentially arbitrary (ties go to the
  alphabetically first language):

  ```{r}
  classify_text(c("Python programming", "John Smith"))
  ```

- **Short texts are unreliable.** The fewer stopwords, the closer every
  language scores to the uniform prior (1/57 ≈ 0.018), so single words and
  greetings are effectively random guesses.

- **Closely related languages are confused.** Portuguese/Spanish,
  Danish/Norwegian/Swedish and Czech/Slovak share many stopwords; the
  winner is often right but with low confidence, with sibling languages
  close behind.

- **Unknown languages are forced into a known one.** A text in a language
  missing from the `stopwords` source (e.g. Icelandic or Amharic) gets an
  arbitrary label at near-uniform probability:

  ```{r}
  classify_text(c("Ég heiti Jón og ég bý í Reykjavík",  # Icelandic, not covered
                  "ሰላም እንዴት ነህ ስሜ አሊሰን ነው"))       # Amharic, not covered
  ```

- **Mixed-language texts get a single label**, usually the language
  contributing the most stopwords.

- **Script and spelling matter.** Stopwords are matched in their native
  script, and text is lowercased but not diacritic-folded, so
  transliterated or accented variants may not match.

- **Probabilities are not calibrated.** The posterior probabilities rank
  the candidates but do not measure real-world accuracy; use `threshold`
  as a coarse safety valve (see [Handling low-confidence
  predictions](#handling-low-confidence-predictions)).

For robust, production-grade detection, consider dedicated tools such as
[fastText](https://fasttext.cc/), [CLD3](https://github.com/google/cld3) or
[langdetect](https://pypi.org/project/langdetect/).

## Where to go next

- `?build_classifier` — train a model from any `stopwords` source.
- `?available_languages` — list languages for a given source.
- `?classify_text` — full reference for the classification function.
